---
title: "A tutorial to select participants and create two balanced groups using propensity scores"
author: "Janet Bang, Megha Sharda, and Aparna Nadig"
date: "February 9, 2020"
output:
  html_document:
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE)

knitr::opts_chunk$set(fig.width = 11, fig.height = 7) 
```

### Load libraries and set theme
```{r}
library(tidyverse)
library(MatchIt)
library(GGally)
library(compute.es)
library(compareGroups)
library(gridExtra)

theme_set(theme_bw())
```


### Import files with data
```{r}
demo <- read_csv("./data/demographics.csv")
celf <- read_csv("./data/celf.csv")
leiter <- read_csv("./data/leiter.csv")
scq <- read_csv("./data/scq.csv")
vineland <- read_csv("./data/vineland.csv")
```




This code follows four main steps to select participants for matched groups. A discussion of matching and a tutorial of these steps can be seen in Bang, Sharda, and Nadig (in preparation):

 Step 1: Assess data

 Step 2: Select covariates

 Step 3: Conduct matching

 Step 4: Diagnose matching

 NOTE: Steps 3-4 are iterated until matching is satisfactory
 
 
 
# Step 1: Assess data

Five TD children and three children from the ASD group were tested but excluded from the study. The final possible sample size to include is 43 TD children and 25 children with ASD. 

### Select participants to include for group matching and combine data to one dataframe
```{r}
df <- demo %>% 
  filter(keep == 1) %>% # final children to include
  left_join(celf, by = "participant") %>% 
  left_join(leiter, by = "participant") %>% 
  left_join(scq, by = "participant") %>% 
  left_join(vineland, by = "participant")

df %>% group_by(group) %>% count()
```





# Step 2: Select covariates

Prior to data collection, we considered multiple covariates that could influence children’s performance on the experimental tasks. Ultimately, we could not include all possible covariates given our sample size, thus we decided on our final two covariates of age and IQ. 

Initial covariates under consideration were based on their known relationships with referential gaze following and word learning in experimental tasks. These covariates included: age, nonverbal IQ, language ability, sex, and parental education. We decided to exclude covariates of sex and parental education. Though both variables are known to be related to broader language abilities in the general population, we prioritized age, nonverbal IQ, and language ability due to their known relations with referential gaze following and word learning in children with ASD.

Given the differences seen in language abilities (particularly on Recalling Sentences and Word Classes), we determined that matching groups on one or more language measures in addition to age and IQ would likely result in excluding some of our 25 children with ASD. Thus, in order to retain as many children as possible, we chose not to include language abilities as a covariate in our matching process. 


### Select participant variables
```{r}
df2 <- df %>% 
  select(participant, keep, group, age_yrs, language, sex, block1, parental_ed, 
         leiter_composite, celf4_RS_scaled, celf4_WC_rec_scaled, celf4_WC_exp_scaled, celf4_WC_total_scaled, 
         celf4_WA_total, scq_total, social_standScore)
```


### Create matrix of covariates to consider for propensity scores
```{r}
mtx <- df2 %>% 
  select("age_yrs", "leiter_composite", "celf4_RS_scaled", "celf4_WC_total_scaled", "celf4_WA_total", "group") %>% 
  rename("Age (yr)" = age_yrs, "Leiter (IQ)" = leiter_composite, "CELF4-RS" = celf4_RS_scaled, 
         "CELF4-WC" = celf4_WC_total_scaled, "CELF4-WA" = celf4_WA_total, "Group" = group)

# Age: years
# Leiter: Nonverbal IQ, composite score (normed: M = 100, SD = 15)
# CELF4 - RS: CELF-4 Recalling Sentences, scaled score (normed: M = 10, SD = 3)
# CELF4 - WC: CELF-4 Word Classes Total (Receptive and Expressive), scaled score (normed: M = 10, SD = 3)
# CELF4 - WA: CELF-4 Word Associations, total score
corr_mtx <- ggpairs(data = mtx, 
        columns = 1:6, 
        title = "Correlation Matrix of Potential Covariates", 
        mapping = aes(colour = Group, alpha = 1),
        axisLabels = "show") + 
        theme(text= element_text(size = 15))

corr_mtx
ggsave("./plots/corr_mtx.pdf", height = 8, width = 11, units = "in")
```


### Descriptive statistics of IQ and age before matching
```{r}
# subset only age and IQ
df2_age_IQ <- df2 %>% 
  select(group, age_yrs, leiter_composite)

# compare groups
df2_age_IQ_compare <- compareGroups(group ~ age_yrs + leiter_composite , data = df2_age_IQ)
df2_age_IQ_compare_tbl <- createTable(df2_age_IQ_compare, digits = 2, type = 1, show.n = TRUE)
print(df2_age_IQ_compare_tbl, header.labels = c(p.overall = "p-value"))

```







# Step 3: Conduct matching

We used the MatchIt package (Ho et al., 2011) in R version 3.5.0 (R Core Team, 2018) to apply both the nearest neighbor and optimal matching algorithms. Both algorithms resulted in the same 25 typically-developing children chosen as matches to the 25 children with ASD. 

### Create dataframe with covariates to be used for propensity scores
```{r}
df_ps_n68 <- df2 %>% 
  select(participant, group, age_yrs, leiter_composite) %>% 
  mutate(group.int = ifelse(group == "ASD", 1, 0)) # group must be an integer for MatchIt
```


### Nearest neighbor n = 68
```{r}
# apply matching
m_near_n68 <- matchit(group.int ~ age_yrs + leiter_composite, 
                     data = df_ps_n68, method = "nearest")

# summary
summary(m_near_n68, data = df_ps_n68)

# obtain list of matched participants
m_near_n68_df <- match.data(m_near_n68)
head(m_near_n68_df)
m_near_n68_df$participant
```


### Optimal matching n = 68
```{r}
# apply matching
m_opt_n68 <- matchit(group.int ~ age_yrs + leiter_composite, 
                     data = df_ps_n68, method = "optimal")

# summary
summary(m_opt_n68, data = df_ps_n68)

# obtain list of matched participants
m_opt_n68_df <- match.data(m_opt_n68)
head(m_opt_n68_df)
m_opt_n68_df$participant
```


### Check to see participants selected in nearest neighbor and optimal matching methods
```{r}
m_near_n68_part <- m_near_n68_df$participant
m_opt_n68_part <- m_opt_n68_df$participant

check_n68_df <- cbind(m_near_n68_part, m_opt_n68_part)
check_n68_df
```


# Step 4: Diagnose matching

### Propensity scores distribution
We first examined how well groups were matched on their propensity scores. A visual inspection of the propensity score plot depicted the same ASD participant with a high propensity score without a close match in the TD group. This score meant that this participant had a high conditional probability of being assigned to the ASD group given the child’s age and IQ. We next examined matching of propensity scores based on cutoff values in the literature: standardized mean difference (d) close to an absolute difference of .25 and variance ratios within .5 to 2 (vr). As seen in the table below, Cohen's d was close to .25 although the vr was within the acceptable range (d = .24, vr = 1.46). Given the high standardized mean difference and the outlier seen in the propensity score plot, we removed the outlier ASD participant and conducted nearest neighbor and optimal matching methods with a revised sample of 24 children with ASD and 43 TD children.
```{r}
# Nearest neighbor (the same participant is identified using the optimal matching method)
plot(m_near_n68, type = "jitter")

# participant with high propensity score in ASD group
m_near_n68_df %>% filter(distance > .7)
```


### Standardized mean difference and variance ratios
```{r}
# NOTE: both methods have the same participants, so either method will provide the same scores. Both are presented here for clarity.
# Nearest neighbor
# a) create vectors with data in asd and td group
near_asd_n68 <- m_near_n68_df %>% 
  filter(group == "ASD")

near_td_n68 <- m_near_n68_df %>% 
  filter(group == "TD")

# b) standardized mean difference of propensity score
# (Mtreatment - Mcontrol) / SDtreatment; (Stuart, 2010; Rubin, 2001; Rosenbaum & Rubin, 1985 - The American Statistician)
(mean(near_asd_n68$distance) - mean(near_td_n68$distance)) / sd(near_asd_n68$distance)

# c) variance ratio
(sd(near_asd_n68$distance)^2) / (sd(near_td_n68$distance)^2)




# Optimal matching
# a) create vectors with data in asd and td group
opt_asd_n68 <- m_opt_n68_df %>% 
  filter(group == "ASD")

opt_td_n68 <- m_opt_n68_df %>% 
  filter(group == "TD")

# b) standardized mean difference of propensity score
(mean(opt_asd_n68$distance) - mean(opt_td_n68$distance)) / sd(opt_asd_n68$distance)

# c) variance ratio
(sd(opt_asd_n68$distance)^2) / (sd(opt_td_n68$distance)^2)
```



# Step 3 with revised sample (24 ASD, 42 TD): Conduct matching
The second iteration using the revised sample resulted in 24 children with ASD matched to 24 TD children. Different children were selected for matched groups when using the nearest neighbor or optimal matching. Propensity score distributions did not indicate clear outliers with either method. However, an examination of standardized mean differences and variance ratios indicated that the optimal matching method was better than the nearest neighbor results (optimal: d = .14, vr = 1.04; nearest neighbor: d = .24, vr = 1.32). These findings suggest that the optimal matching method with 24 children per group resulted in balanced groups.  

### Remove participant in ASD group with high propensity score
```{r}
df_ps_n67 <- df2 %>% 
  select(participant, group, age_yrs, leiter_composite) %>% 
  mutate(group.int = ifelse(group == "ASD", 1, 0)) %>% 
  filter(participant != 330)

# check count of participants ASD = 24 and TD = 43
df_ps_n67 %>% count(group)
```


### Nearest neighbor n = 67
```{r}
# apply matching
m_near_n67 <- matchit(group.int ~ age_yrs + leiter_composite, 
                     data = df_ps_n67, method = "nearest")

# summary
summary(m_near_n67, data = df_ps_n67)

# obtain list of matched participants
m_near_n67_df <- match.data(m_near_n67)
head(m_near_n67_df)
m_near_n67_df$participant
```


### Optimal matching n = 67
```{r}
# 1) Run logistic regressions
# runs logistic regression to create matched sample
m_opt_n67 <- matchit(group.int ~ age_yrs + leiter_composite, 
                     data = df_ps_n67, method = "optimal")

# summary
summary(m_opt_n67, data = df_ps_n67)

# obtain list of matched participants
m_opt_n67_df <- match.data(m_opt_n67)
head(m_opt_n67_df)
m_opt_n67_df$participant
```


# Step 4 with revised sample (24 ASD, 42 TD): Diagnose matching

### Propensity scores distribution
```{r}
# Nearest neighbor
plot(m_near_n67, type = "jitter")

# Optimal matching
plot(m_opt_n67, type = "jitter")

```


### Standardized mean difference and variance ratios
```{r}
# Nearest neighbor
# a) create vectors with data in asd and td group
near_asd_n67 <- m_near_n67_df %>% 
  filter(group == "ASD")

near_td_n67 <- m_near_n67_df %>% 
  filter(group == "TD")

# b) standardized mean difference of propensity score
(mean(near_asd_n67$distance) - mean(near_td_n67$distance)) / sd(near_asd_n67$distance)

# c) variance ratio
(sd(near_asd_n67$distance)^2) / (sd(near_td_n67$distance)^2)




# Optimal matching
# a) create vectors with data in asd and td group
opt_asd_n67 <- m_opt_n67_df %>% 
  filter(group == "ASD")

opt_td_n67 <- m_opt_n67_df %>% 
  filter(group == "TD")

# b) standardized mean difference of propensity score
(mean(opt_asd_n67$distance) - mean(opt_td_n67$distance)) / sd(opt_asd_n67$distance)

# c) variance ratio
(sd(opt_asd_n67$distance)^2) / (sd(opt_td_n67$distance)^2)
```

The final step in diagnosing groups is to determine how well groups are matched on each covariate included in the propensity score, as well as any other covariates considered of interest in the study (Blackford, 2007; Ho, Imai, King, & Stuart, 2007; Stuart, 2010). Prior to determining propensity scores, the only variable where we expected group differences was on language variables due to their already different distributions. Guidelines to evaluate well matched groups on each covariate (e.g., nonverbal IQ, age) included examination of boxplots, p values > .5 (Kover & Atwood, 2013; Mervis & Robinson, 1999; Mervis & Klein-Tasman, 2004), Cohen’s d close to 0, and variance ratios close to 1. Cohen’s d was calculated with formulas in line with Kover and Atwood (2013). The use of Cohen’s d and variance ratios are recommended as alternatives to inferential statistics such as p values, due to difficulties with establishing equivalence with inferential statistics (Kover & Atwood, 2013). 


As seen in the table below, our final revised sample with the optimal matching method resulted in two groups successfully balanced according to criteria listed above in their distribution for our covariates of interest, IQ and age. We next examined other variables not included in our propensity score (29). The matched groups met the cut-off for p values > .5 on the ratio of English- to French-speaking children. On measures of sex, maternal education level, and CELF-4 Word Association, groups were not significantly different but did not meet recommended cutoffs of values noted above. As expected prior to selecting participants, groups were significantly different in their distribution on language measures of Recalling Sentences and Word Classes. Additionally, as expected due to diagnoses, groups were significantly different on social skills measures of the SCQ and VABS-II.

Lastly, we verified the distribution of children for randomized factors (i.e., block order). As seen in the table above including descriptive statistics, the same proportion of children had both block orders. In sum, the optimal method with 24 ASD, 24 TD established balanced groups on desired covariates of age and IQ.


### Get list of matched participants and all covariates to report in study
```{r}
matchit_list <- m_opt_n67_df %>% 
  select(participant)

df_matchit <- df2 %>% 
  right_join(matchit_list, by = "participant") %>% 
  select(-c(keep, celf4_WC_rec_scaled, celf4_WC_exp_scaled)) %>% 
  mutate(group = factor(group), language = factor(language), 
         sex = factor(sex), parental_ed = factor(parental_ed), 
         block1 = factor(block1))

nrow(df_matchit)
```


### Violin plots to examine continuous variables
```{r}
p1 <- ggplot(df_matchit, aes(group, age_yrs, fill = group)) + 
  geom_violin(alpha = .7) + 
  geom_jitter(alpha = .3) +
  theme(legend.position = "none", axis.text=element_text(size=15)) + 
  labs(y = "Age (yrs)", x = "")
p2 <- ggplot(df_matchit, aes(group, leiter_composite, fill = group)) + 
  geom_violin(alpha = .7) + 
  geom_jitter(alpha = .3) +
  theme(legend.position = "none", axis.text=element_text(size=15)) + 
  labs(y = "Nonverbal IQ (Leiter)", x = "")
p3 <- ggplot(df_matchit, aes(group, celf4_WA_total, fill = group)) + 
  geom_violin(alpha = .7) + 
  geom_jitter(alpha = .3) +
  theme(legend.position = "none", axis.text=element_text(size=15)) + 
  labs(y = "CELF-4 Word Associations", x = "")

p4 <- ggplot(df_matchit, aes(group, celf4_RS_scaled, fill = group)) + 
  geom_violin(alpha = .7) + 
  geom_jitter(alpha = .3) +
  theme(legend.position = "none", axis.text=element_text(size=15)) + 
  labs(y = "CELF-4 Recalling Sentences", x = "")
p5 <- ggplot(df_matchit, aes(group, celf4_WC_total_scaled, fill = group)) + 
  geom_violin(alpha = .7) + 
  geom_jitter(alpha = .3) +
  theme(legend.position = "none", axis.text=element_text(size=15)) + 
  labs(y = "CELF-4 Word Classes", x = "")
p6 <- ggplot(df_matchit, aes(group, scq_total, fill = group)) + 
  geom_violin(alpha = .7) + 
  geom_jitter(alpha = .3) +
  theme(legend.position = "none", axis.text=element_text(size=15)) + 
  labs(y = "Social Communication Questionnaire", x = "")
p7 <- ggplot(df_matchit, aes(group, social_standScore, fill = group)) + 
  geom_violin(alpha = .7) +
  geom_jitter(alpha = .3) +
  theme(legend.position = "none", axis.text=element_text(size=15)) + 
  labs(y = "Vineland Socialization Domain", x = "")

grid.arrange(p1, p2, p3, p4, p5, p6, p7, nrow = 2)
```


### Inferential statistics
```{r}
# remove participant column to run compareGroups
df_matchit2 <- df_matchit %>% 
  select(-participant)

# compareGroups
# compareGroups uses Fisher's exact test as a default when cell values < 5
compare_group <- compareGroups(group ~ . , data = df_matchit2)
compare_group_tbl <- createTable(compare_group, digits = 2, type = 1, show.n = TRUE)
print(compare_group_tbl, header.labels = c(p.overall = "p-value"))
```


### Cohen's d and variance ratios
```{r}
# create separate vectors for groups
asd <- df_matchit %>% 
  filter(group == "ASD")

td <- df_matchit %>% 
  filter(group == "TD")

# age_yrs - cohen's d and variance ratios
age_yrs_es <- mes(mean(asd$age_yrs), mean(td$age_yrs),
                  sd(asd$age_yrs), sd(td$age_yrs),
                  nrow(asd), nrow(td), verbose = F)
age_yrs <- age_yrs_es[, 4]
age_yrs_es_ci <- age_yrs_es[, 6:7]
age_yrs_es_vr <- (sd(asd$age_yrs)^2)/(sd(td$age_yrs)^2)


# celf4_RS_scaled - cohen's d and variance ratios
celf4_rs_es <- mes(mean(asd$celf4_RS_scaled), mean(td$celf4_RS_scaled),
                   sd(asd$celf4_RS_scaled), sd(td$celf4_RS_scaled),
                   nrow(asd), nrow(td), verbose = F)
celf4_rs <- celf4_rs_es[, 4]
celf4_rs_es_ci <- celf4_rs_es[, 6:7]
celf4_rs_es_vr <- (sd(asd$celf4_RS_scaled)^2)/(sd(td$celf4_RS_scaled)^2)


# celf4_WC_total_scaled - cohen's d and variance ratios
# removing child with no score for WC
asd2 <- asd %>% 
  filter(participant != 420)

celf4_wc_es <- mes(mean(asd2$celf4_WC_total_scaled), mean(td$celf4_WC_total_scaled),
                   sd(asd2$celf4_WC_total_scaled), sd(td$celf4_WC_total_scaled),
                   nrow(asd2), nrow(td), verbose = F)
celf4_wc <- celf4_wc_es[, 4]
celf4_wc_es_ci <- celf4_wc_es[, 6:7]
celf4_wc_es_vr <- (sd(asd2$celf4_WC_total_scaled)^2)/(sd(td$celf4_WC_total_scaled)^2)


# celf4_WA_total - cohen's d and variance ratios
celf4_wa_es <- mes(mean(asd$celf4_WA_total), mean(td$celf4_WA_total),
                   sd(asd$celf4_WA_total), sd(td$celf4_WA_total),
                   nrow(asd), nrow(td), verbose = F)
celf4_wa <- celf4_wa_es[, 4]
celf4_wa_es_ci <- celf4_wa_es[, 6:7]
celf4_wa_es_vr <- (sd(asd$celf4_WA_total)^2)/(sd(td$celf4_WA_total)^2)


# leiter_composite - cohen's d and variance ratios
leiter_es <- mes(mean(asd$leiter_composite), mean(td$leiter_composite),
                 sd(asd$leiter_composite), sd(td$leiter_composite),
                 nrow(asd), nrow(td), verbose = F)
leiter <- leiter_es[, 4]
leiter_es_ci <- leiter_es[, 6:7]
leiter_es_vr <- (sd(asd$leiter_composite)^2)/(sd(td$leiter_composite)^2)


# scq_total - cohen's d and variance ratios
scq_es <- mes(mean(asd$scq_total), mean(td$scq_total),
              sd(asd$scq_total), sd(td$scq_total),
              nrow(asd), nrow(td), verbose = F)
scq <- scq_es[, 4]
scq_es_ci <- scq_es[, 6:7]
scq_es_vr <- (sd(asd$scq_total)^2)/(sd(td$scq_total)^2)


# social_standScore - cohen's d and variance ratios
vineland_es <- mes(mean(asd$social_standScore), mean(td$social_standScore),
                   sd(asd$social_standScore), sd(td$social_standScore),
                   nrow(asd), nrow(td), verbose = F)
vineland <- vineland_es[, 4]
vineland_es_ci <- vineland_es[, 6:7]
vineland_es_vr <- (sd(asd$social_standScore)^2)/(sd(td$social_standScore)^2)


es_d <- rbind(age_yrs, celf4_rs, celf4_wc, celf4_wa, leiter, scq, vineland)
es_ci <- rbind(age_yrs_es_ci, celf4_rs_es_ci, celf4_wc_es_ci, celf4_wa_es_ci, leiter_es_ci, scq_es_ci, vineland_es_ci)
es_vr <- rbind(age_yrs_es_vr, celf4_rs_es_vr, celf4_wc_es_vr, celf4_wa_es_vr, leiter_es_vr, scq_es_vr, vineland_es_vr)

effect_sizes <- cbind(es_d, es_ci, es_vr)
effect_sizes %>% 
  rename( d = 1, d_CI_lower_bound = 2, d_CI_upper_bound = 3, vr = 4)

```



# Examine matched groups when including age, nonverbal IQ, and language (Recalling Sentences)

We explored matches when including three covariates of age, nonverbal IQ, and language. As seen below, both nearest neighbor and optimal matching resulted in selecting different potential matches of TD children, but there were still 8 children with ASD whose propensity scores were outside the range of all typically-developing children. An examination of balancing on propensity scores and individual covariates indicates that only age meets the desired cutoff of p > .5, and the distribution of CELF-4 Recalling Sentences appears substantially different between both groups (ps < .5 for two variables, propensity score Cohen’s d > .5 and propensity score variance ratios > 3). 

Due to the poor balancing when including all three proposed covariates, this evidence supports balancing on two covariates of age and nonverbal IQ. Our visual analysis of language measures prior to data analyses confirms these post-hoc exploratory analyses. We initially chose to match only on age and nonverbal IQ because we wanted to retain as many children in the sample as possible. Additionally, because language abilities of children with ASD were not categorically poorer across all three measures relative to TD children, it was unclear which language measure to match on when intercorrelations between language measures ranged widely across the full sample (rs = .28 - .72), within ASD (rs = .34 - .84) and within TD children (rs = .09 - .47). Therefore, matching groups on covariates of age and nonverbal IQ appear to be both theoretically supported based on prior studies and empirically supported by the current evidence with our sample.


### Select covariates
```{r}
df_ps_n68_3cov <- df2 %>% 
  select(participant, group, age_yrs, leiter_composite, celf4_RS_scaled) %>% 
  mutate(group.int = ifelse(group == "ASD", 1, 0)) # group must be an integer for MatchIt
```


### Nearest neighbor n = 68 - including age, IQ, and language (Recalling Sentences)
```{r}
# apply matching
m_near_n68_3cov <- matchit(group.int ~ age_yrs + leiter_composite + celf4_RS_scaled, 
                     data = df_ps_n68_3cov, method = "nearest")

# summary
summary(m_near_n68_3cov, data = df_ps_n68_3cov)

# obtain list of matched participants
m_near_n68_3cov_df <- match.data(m_near_n68_3cov)
head(m_near_n68_3cov_df)
m_near_n68_3cov_df$participant
```


#### Diagnose matching - View propensity scores
```{r}
# View propensity scores
plot(m_near_n68_3cov, type = "jitter")
```


#### Diagnose matching - Standardized mean difference and Variance ratio
```{r}
# Standardized mean difference
# a) create vectors with data in asd and td group
near_asd_n68_3cov <- m_near_n68_3cov_df %>% 
  filter(group == "ASD")

near_td_n68_3cov <- m_near_n68_3cov_df %>% 
  filter(group == "TD")

# b) standardized mean difference of propensity score
# (Mtreatment - Mcontrol) / SDtreatment; (Stuart, 2010; Rubin, 2001; Rosenbaum & Rubin, 1985 - The American Statistician)
(mean(near_asd_n68_3cov$distance) - mean(near_td_n68_3cov$distance)) / sd(near_asd_n68_3cov$distance)

# Variance ratio
(sd(near_asd_n68_3cov$distance)^2) / (sd(near_td_n68_3cov$distance)^2)
```


#### Diagnose matching - Descriptive and inferential statistics on age, IQ, and language
```{r}
# subset only age and IQ
age_IQ_language_3cov <- m_near_n68_3cov_df %>% 
  select(group, age_yrs, leiter_composite, celf4_RS_scaled)

# compare groups
age_IQ_language_compare <- compareGroups(group ~ age_yrs + leiter_composite + celf4_RS_scaled, data = age_IQ_language_3cov)
age_IQ_language_compare_tbl <- createTable(age_IQ_language_compare, digits = 2, type = 1, show.n = TRUE)
print(age_IQ_language_compare_tbl, header.labels = c(p.overall = "p-value"))

```




### Optimal matching n = 68 - including age, IQ, and language (Recalling Sentences)
```{r}
# apply matching
m_opt_n68_3cov <- matchit(group.int ~ age_yrs + leiter_composite + celf4_RS_scaled, 
                     data = df_ps_n68_3cov, method = "optimal")

# summary
summary(m_opt_n68_3cov, data = df_ps_n68_3cov)

# obtain list of matched participants
m_opt_n68_3cov_df <- match.data(m_opt_n68_3cov)
head(m_opt_n68_3cov_df)
m_opt_n68_3cov_df$participant
```


#### Diagnose matching - View propensity scores
```{r}
# View propensity scores
plot(m_opt_n68_3cov, type = "jitter")
```


#### Diagnose matching - Standardized mean difference and Variance ratio
```{r}
# Standardized mean difference
# a) create vectors with data in asd and td group
opt_asd_n68_3cov <- m_opt_n68_3cov_df %>% 
  filter(group == "ASD")

opt_td_n68_3cov <- m_opt_n68_3cov_df %>% 
  filter(group == "TD")

# b) standardized mean difference of propensity score
# (Mtreatment - Mcontrol) / SDtreatment; (Stuart, 2010; Rubin, 2001; Rosenbaum & Rubin, 1985 - The American Statistician)
(mean(opt_asd_n68_3cov$distance) - mean(opt_td_n68_3cov$distance)) / sd(opt_asd_n68_3cov$distance)

# Variance ratio
(sd(opt_asd_n68_3cov$distance)^2) / (sd(opt_td_n68_3cov$distance)^2)
```


#### Diagnose matching - Descriptive and inferential statistics on age, IQ, and language
```{r}
# subset only age and IQ
age_IQ_language_3cov <- m_opt_n68_3cov_df %>% 
  select(group, age_yrs, leiter_composite, celf4_RS_scaled)

# compare groups
age_IQ_language_compare <- compareGroups(group ~ age_yrs + leiter_composite + celf4_RS_scaled, data = age_IQ_language_3cov)
age_IQ_language_compare_tbl <- createTable(age_IQ_language_compare, digits = 2, type = 1, show.n = TRUE)
print(age_IQ_language_compare_tbl, header.labels = c(p.overall = "p-value"))

```

### Diagnose matching - Descriptive and inferential statistics on age, IQ, and language - removing 8 ASD outliers
```{r}
# subset only age and IQ
age_IQ_language_3cov_rm8asd <- m_opt_n68_3cov_df %>% 
  filter(distance < .7) %>% 
  select(group, age_yrs, leiter_composite, celf4_RS_scaled)
  

# compare groups
age_IQ_language_compare_rm8asd <- compareGroups(group ~ age_yrs + leiter_composite + celf4_RS_scaled, data = age_IQ_language_3cov_rm8asd)
age_IQ_language_compare_rm8asd_tbl <- createTable(age_IQ_language_compare_rm8asd, digits = 2, type = 1, show.n = TRUE)
print(age_IQ_language_compare_rm8asd_tbl, header.labels = c(p.overall = "p-value"))

```

